---
title: "Caching in R"
author: "Oleksandr Fialko"
date: "11/22/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Matrix inversion is usually a costly computation and there may be some benefit to caching the inverse of a matrix rather than compute it repeatedly. Below is a pair of functions that cache the inverse of a matrix.

### Function 1
Function named `makeCacheMatrix` creates a special "matrix" object that can cache its inverse. This function creates a list object from the input matrix containing functions to
   1  set the value of the matrix
   2  get the value of the matrix
   3  set the value of the inverse
   4  get the value of the inverse

```{r echo=TRUE,cache=TRUE}
makeCacheMatrix <- function(x = matrix()) {
    inv <- NULL
    set <- function(y) {
        x <<- y
        inv <<- NULL
    }
    get <- function() x
    setinv <- function(inverse) inv <<- inverse
    getinv <- function() inv
    list(set = set, get = get,
         setinv = setinv,
         getinv = getinv)
}
```


### Function 2
Function named `cacheSolve` computes the inverse of the special "matrix" returned by `makeCacheMatrix` above. If the inverse has already been calculated (and the matrix has not changed), then the `cachesolve` retrieves the inverse from the cache. Computing the inverse of a square matrix can be done with the solve function in R. 

```{r echo=TRUE,cache=TRUE}
cacheSolve <- function(x, ...) {
    inv <- x$getinv()
    if(!is.null(inv)) {
        message("getting cached data")
        return(inv)
    }
    data <- x$get()
    inv <- solve(data, ...)
    x$setinv(inv)
    inv
}
```


### Testing

Create a matrix:
```{r echo=TRUE,cache=TRUE}
tridiag <- function(upper, lower, main){
    out <- matrix(0,length(main),length(main))
    diag(out) <- main
    indx <- seq.int(length(upper))
    out[cbind(indx+1,indx)] <- lower
    out[cbind(indx,indx+1)] <- upper
    return(out)
}

h <- tridiag(rep(1,999),rep(1,999),rep(2,1000))
hc <- makeCacheMatrix(h)
```

First call:
```{r echo=TRUE,cache=TRUE}
system.time(cacheSolve(hc))
```

Second call:
```{r echo=TRUE,cache=TRUE}
system.time(cacheSolve(hc))
```

