---
title: "Getting and Cleaning Human Activity Recognition Using Smartphones Data Set "
author: "Oleksandr Fialko"
date: "12/7/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

One of the most exciting areas in all of data science right now is wearable computing. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available [at the site](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) where the data was obtained.

The goal here is to prepare tidy data that can be used for later analysis.

## Step 1: Merging the training and the test sets to create one data set

Creating data directory to store data
```{r, eval=FALSE}
if (!file.exists('data')){
  dir.create('data')
}
```

Downloading data
```{r eval=FALSE}
fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileURL,destfile = 'data/data_project.zip')
unzip('data/data_project.zip',exdir = 'data/')
date_of_download <- date()
```

Uploading the files into R
```{r eval=FALSE}
features <- read.table('data/UCI HAR Dataset/features.txt',
                       col.names = c('Index_Feature','Feature'))
activity_labels <- read.table('data/UCI HAR Dataset/activity_labels.txt',
                              col.names = c('Index_Activity','Activity'))

X_train <- read.table('data/UCI HAR Dataset/train/X_train.txt',
                      col.names = features$Feature,check.names = FALSE)
y_train <- read.table('data/UCI HAR Dataset/train/y_train.txt')
subject_train <- read.table('data/UCI HAR Dataset/train/subject_train.txt')

X_test <- read.table('data/UCI HAR Dataset/test/X_test.txt',
                     col.names = features$Feature,check.names = FALSE)
y_test <- read.table('data/UCI HAR Dataset/test/y_test.txt')
subject_test <- read.table('data/UCI HAR Dataset/test/subject_test.txt')

X_train_p = cbind(y_train, subject_train,X_train)
X_test_p = cbind(y_test, subject_test,X_test)
```

Renaming first two columns
```{r eval=FALSE}
names(X_train_p)[c(1,2)] <- c('Activity','Subject')
names(X_test_p)[c(1,2)] <- c('Activity','Subject')
```

Storring transformed data
```{r eval=FALSE}
write.csv(x=X_train_p,file = 'X_train.csv',sep = ',',row.names = F)
write.csv(x=X_test_p,file = 'X_test.csv',sep = ',',row.names = F)
```

Merging train and test sets
```{r eval=FALSE}
X_data <- rbind(X_train,X_test)
y_data <- rbind(y_train,y_test)
subject_data <- rbind(subject_train,subject_test)
```

## Step 2: Extracting only the measurements on the mean and standard deviation

```{r eval=FALSE}
col_names <- colnames(X_data)
col_mean  <- grep('mean\\(\\)',col_names)
col_std   <- grep('std\\(\\)',col_names)

X_mean_std <- X_data[,c(col_mean,col_std)]
```


## Step 3: Using descriptive activity names

```{r eval=FALSE}
X_clean <- cbind(y_data,X_mean_std)
colnames(X_clean)[1] <- 'Index_Activity'
X_clean <- cbind(subject_data,X_clean)
colnames(X_clean)[1] <- 'Subject'

X_clean <- merge(activity_labels,X_clean)
X_clean$Index_Activity <- NULL
```

## Step 4: Appropriatelly labels the data set with descriptive variable names

```{r eval=FALSE}
col_names <- colnames(X_clean)

col_names <- sub(x = col_names,pattern = '^t',replacement = 'Time domain signal: ')
col_names <- sub(x = col_names,pattern = '^f',replacement = 'Frequency domain signal: ')
col_names <- sub(x = col_names,pattern = '-',replacement = ', ')
col_names <- sub(x = col_names,pattern = 'mean\\(\\)',replacement = ' mean value ')
col_names <- sub(x = col_names,pattern = 'std\\(\\)',replacement = ' standart deviation ')
col_names <- sub(x = col_names,pattern = '-X',replacement = 'in X direction')
col_names <- sub(x = col_names,pattern = '-Y',replacement = 'in Y direction')
col_names <- sub(x = col_names,pattern = '-Z',replacement = 'in Z direction')
col_names <- sub(x = col_names,pattern = 'AccJerk',replacement = ' acceleration jerk')
col_names <- sub(x = col_names,pattern = 'Acc',replacement = ' acceleration')
col_names <- sub(x = col_names,pattern = 'GyroJerk',replacement = ' angular velocity jerk')
col_names <- sub(x = col_names,pattern = 'Gyro',replacement = ' angular velocity')
col_names <- sub(x = col_names,pattern = 'Mag',replacement = ' magnitude')

colnames(X_clean) <- col_names
```

## Step 5: Tidy data set with the average of each variable for each activity and each subject

```{r eval=FALSE}
X_tidy <- aggregate(X_clean[,3:68],by=list(X_clean$Activity,X_clean$Subject),FUN=mean)
colnames(X_tidy)[1] <- 'Activity'
colnames(X_tidy)[2] <- 'Subject'
```

## Step 6: Storing the final tidy data set

```{r eval=FALSE}
write.table(X_tidy,file = 'tidy_data_set.txt',row.names = F)
```

